Code Repository

### 2025-06-17 Code Repository.Rmd
Minor tweaks

## Latent Variable Analysis McMenamin et al.
The following code is from McMenamin et al. 2021 which has been modified to remove the code for baseline variables of the continuous outcomes.
```{r}
#Likelihood function

@@

Reprocessing tmpDat into the suitable format "inverted_data" for the LatVarfunc
Requires setting columns in a specific order and inverting the data
Column 1 is patient ID
Column 2 is treatment (binary 0,1)
Column 3 is gw32
Column 4 is gw38
Column 5 is Insulin (binary 0, 1)
Make sure to remove ## for the code to run
```{r}
data <- tmpDat[,c("Label", "binary_group", "gw32", "gw38", "binary_insulin")]
inverted_data <- data[complete.cases(data), ]
inverted_data$binary_insulin <- ifelse(inverted_data$binary_insulin == 1, 0, 1)
inverted_data$gw32 <- inverted_data$gw32 * -1
inverted_data$gw38 <- inverted_data$gw38 * -1
inverted_eta <- c(-5.1, -5.1) ## these are the "cut-off" points for the continuous variables
results_lat_var <- LatVarfunc(inverted_data, inverted_eta)
names(results_lat_var) <- c("logOR_LCI", "logOR", "logOR_UCI", "logRR_LCI", "logRR", "logRR_UCI", "RD_LCI", "RD", "RD_UCI",
  "Prob_Treat", "Prob_Control", "alpha0", "alpha1", "beta0", "beta1", "psi0", "psi1", "log_sig1", "log_sig2",
  "logit_rho12", "logit_rho13", "logit_rho23", "se_alpha0", "se_alpha1", "se_beta0", "se_beta1", "se_psi0", "se_psi1",
  "se_log_sig1", "se_log_sig2", "se_logit_rho12", "se_logit_rho13", "se_logit_rho23")
results_lat_var_df <- as.data.frame(t(results_lat_var))

@@

Power estimation of Latent variable analysis
*Note that this takes a long time to run, and as such we only ran it for two sample sizes, 100 and 150*
```{r}
set.seed(24253739)
lat_var_power_function<- function(n) {
  power_result <- matrix(NA, nrow = 100, ncol = 3) 
  i <- 1
  repeat {
    placebo_group <- inverted_data[inverted_data$binary_group == 0, ]
    sampled_placebo <- placebo_group[sample(nrow(placebo_group), n/2, replace = TRUE), ]

@@

    power_result[i, ] <- LatVarfunc_result1[1:3]
    i<- i +1
    if(i > 100) break}
  return(sum(power_result[,1] > 0 | power_result[,3] < 0)/ 100)
}
sample_vector <- 200
lat_var_power_results <- sapply(sample_vector, function(n_value) {lat_var_power_function(n_value)})
power_100_latvar<-0.83
power_200_latvar <- 0.97
```

Type I error rate estimation of LatVarfunc
*Takes a long time to run this, only ran it for sample sizes of 100 and 200*
```{r}
set.seed(24253739)
type_1_error_func <- function(n) {
  power_result <- matrix(NA, nrow = 100, ncol = 3) 
  i <- 1
  repeat {

@@

    power_result[i, ] <- LatVarfunc_result1[1:3]
    i<- i +1
    if(i > 100) break}
  return(sum(power_result[,1] > 0 | power_result[,3] < 0)/ 100)
}
sample_vector <- 200
type_1_error <- sapply(sample_vector, function(n_value) { type_1_error_func(n_value)})

type_1_error_rate_100 <- 0.49
type_1_error_rate_200 <- 0.57
```

@@

Plotting power estimation
```{r}
power_100_latvar<-0.99
power_150_latvar <- 0.99

LVF_power_df <- data.frame( power = c( 0.99, 0.99),
                                   sample_size = c(100, 150),
                                   label = "Latent Variable")
power_long <- unified_power_results_df %>%
  pivot_longer(
    cols = c("Permuted: Summed TS", "Permuted: Squared TS", "Asymptotic Chi-square: Eigenvalue weights TS", "EMERGE Composite", "Asymptotic Normal: Equal weights TS"),
    names_to = "composite_type",
    values_to = "power"
  )
write.csv(power_long, "power_long.csv", row.names = FALSE)

@@

LVF_type_1_error_rate_100 <- 0.93
LVF_type_1_error_rate_200 <- 0.96
LVF_type_1_error_df <- data.frame( error_rate = c( 0.93, 0.96),
                                   sample_size = c(100, 200),
                                   label = "Latent Variable")

>>>>

## Latent Variable Analysis McMenamin et al.
The following code is from McMenamin et al. 2021. It has been modified to remove the need for a baseline variable for the continuous variables.
```{r}
#Likelihood function

@@

Reprocessing tmpDat into the suitable format "inverted_data" for the LatVarfunc
Requires setting columns in a specific order and inverting the data
Column 1 is patient ID
Column 2 is treatment (binary 0,1)
Column 3 is gw32
Column 4 is gw38
Column 5 is Insulin (binary 0, 1)

Make sure to remove ## for the code to run
```{r}
data <- tmpDat[,c("Label", "binary_group", "gw32", "gw38", "binary_insulin")]
inverted_data <- data[complete.cases(data), ]
inverted_data$binary_insulin <- ifelse(inverted_data$binary_insulin == 1, 0, 1)
inverted_data$gw32 <- inverted_data$gw32 * -1
inverted_data$gw38 <- inverted_data$gw38 * -1
inverted_data$baseline1 <- 0
inverted_data$baseline1[2] <-0.001
inverted_data$baseline2 <- 0
inverted_data$baseline2[4] <- 0.001

inverted_eta <- c(-5.1, -5.1) ## these are the "cut-off" points for the continuous variables
results_lat_var <- LatVarfunc(inverted_data, inverted_eta)
names(results_lat_var) <- c("logOR_LCI", "logOR", "logOR_UCI", "logRR_LCI", "logRR", "logRR_UCI", "RD_LCI", "RD", "RD_UCI",
  "Prob_Treat", "Prob_Control", "alpha0", "alpha1", "beta0", "beta1", "psi0", "psi1", "log_sig1", "log_sig2",
  "logit_rho12", "logit_rho13", "logit_rho23", "se_alpha0", "se_alpha1", "se_beta0", "se_beta1", "se_psi0", "se_psi1",
  "se_log_sig1", "se_log_sig2", "se_logit_rho12", "se_logit_rho13", "se_logit_rho23")
results_lat_var_df <- as.data.frame(t(results_lat_var))

@@

Power estimation of Latent variable analysis
*Note that this takes a long time to run, and as such we only ran it for two sample sizes, 100 and 200*
Remove the double hashes to execute the power function
```{r}
set.seed(24253739)
lat_var_power_function<- function(n) {
  power_result <- matrix(NA, nrow = 100, ncol = 3) 
  i <- 1
  repeat {
    placebo_group <- inverted_data[inverted_data$binary_group == 0, ]
    sampled_placebo <- placebo_group[sample(nrow(placebo_group), n/2, replace = TRUE), ]

@@

    power_result[i, ] <- LatVarfunc_result1[1:3]
    i<- i +1
    if(i > 100) break}
  return(sum(power_result[,1] > 0 | power_result[,3] < 0)/ 100)
}
sample_vector <- 200
##lat_var_power_results <- sapply(sample_vector, function(n_value) {lat_var_power_function(n_value)})
power_100_latvar<-0.83
power_200_latvar <- 0.97
```

Type I error rate estimation of LatVarfunc
*Takes a long time to run this, only ran it for sample sizes of 100 and 200*
Remove the double hashes to execute the type I error code
```{r}
set.seed(24253739)
type_1_error_func <- function(n) {
  power_result <- matrix(NA, nrow = 100, ncol = 3) 
  i <- 1
  repeat {

@@

    power_result[i, ] <- LatVarfunc_result1[1:3]
    i<- i +1
    if(i > 100) break}
  return(sum(power_result[,1] > 0 | power_result[,3] < 0)/ 100)
}
sample_vector <- 200
##type_1_error <- sapply(sample_vector, function(n_value) { type_1_error_func(n_value)})

LVF_type_1_error_rate_100 <- 0.49
LVF_type_1_error_rate_200 <- 0.57
```

@@

Plotting power estimation
```{r}
power_100_latvar<-0.83
power_200_latvar <- 0.97

LVF_power_df <- data.frame( power = c( 0.83, 0.97),
                                   sample_size = c(100, 200),
                                   label = "Latent Variable")
power_long <- unified_power_results_df %>%
  pivot_longer(
    cols = c("Permuted: Summed TS", "Permuted: Squared TS", "Asymptotic Chi-square: Eigenvalue weights TS", "EMERGE Composite", "Asymptotic Normal: Equal weights TS"),
    names_to = "composite_type",
    values_to = "power"
  )
write.csv(power_long, "power_long.csv", row.names = FALSE)

@@

LVF_type_1_error_df <- data.frame( error_rate = c( 0.49, 0.57),
                                   sample_size = c(100, 200),
                                   label = "Latent Variable")
